
---
title: "EPIC Pipeline"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  parameterfile:
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
start.Time = Sys.time()
formatted.time = format(Sys.time(), "%Y-%m-%d-%Hh%Mm")
#source("../temporary/parameterfile.R")
source(params$parameterfile)
source("ILMN-part0-split.R")
```
This is a pipeline for preprocessing EPIC-Methylation data using R. 

```{r child = 'prepPart1.Rmd'}
```
```{r child = 'prepPart2.Rmd'}
```
```{r child = 'prepPart3.Rmd'}
```
```{r child = 'prepPart4.Rmd'}
```

## Output

```{r}
analysisfilename = paste0("analysis_ready_",projectname,"_",formatted.time,".Rdata")
qcfilename=paste0("QC_data_",projectname,"_",formatted.time,".Rdata")
```

1. The following data is included in the Rdata-file `r qcfilename`:
\begin{description}
	\item[ctrl.all, ctrl.complete.Red.all, ctrl.complete.Green.all, control.info] control probe data
	\item[ctrlprobes.scores] rotated control probe intensities into the coordinates given by principal component analysis
	\item[dp.all] detection p-values of unfiltered idat data
	\item[TypeII.Red.All etc] list the corresponding intensity values
	\item[TypeII.Red.All.d etc] list the corresponding intensity values where detection p-values smaller than threshold $`r detPthreshold`$ are set to missing (NA).
	\item[sample.call] the sample call rates
	\item[marker.call] the marker call rates
	\item[beta.raw] the $\beta$ values calculated for autosomal probes with detection p value filter only
	\item[beta.raw.sex] the $\beta$ values calculated for sex chromosome probes with detection p value filter only
	\item[pcaAuto] output of function prcomp conducting PCA of \textbf{betaQN}
	\item[sample.callY, sample.callX] sample call rates for chromosomes X and Y for quantile normalized filtered samples
	\item[marker.call.all] marker callrate of all quantile normalized samples incl. the ones excluded by IQR-filter 
	\item[marker.outlier.all] data frame of number of probes, detected outliers and marker callrate after filtering by IQR 
\end{description}


2. The data needed for further analysis is saved as `r analysisfilename`:
\begin{description}
	\item[pcaControls] output of function prcomp conducting the PCA of control probes 
	\item[betaQN] $\beta$ values calculated for autosomal probes of filtered samples with quantile normalization applied 
	\item[betaQN.sex] $\beta$ values calculated for sex chromosome probes of filtered samples with quantile normalization applied 
	\item[est.wbc.minfi] the white-blood-cell estimations by minfi method based on RGsets
	\item[betaQN.all] combines \textbf{betaQN} and \textbf{betaQN.sex} 
	\item[betaQN.all.3IQR] modified \textbf{betaQN.all}: IQR outliers are filtered out
#	\item[mdata] transposed \textbf{betaQN} with Sample Names assigned as rownames
#	\item[mdata.3IQR] transposed \textbf{betaQN.all.3IQR} with Sample Names as rownames
\end{description}

If you run the pipeline a second time the data files will be overwritten, so please store them at a different place if you want to keep both. 

```{r}
save(ctrl.all,ctrlprobes.scores,ctrl.complete.Red.all, ctrl.complete.Green.all, control.info,dp.all, TypeII.Red.All , TypeII.Green.All , TypeI.Red.M.All , TypeI.Red.U.All , TypeI.Green.M.All , TypeI.Green.U.All, TypeII.Red.All.d , TypeII.Green.All.d , TypeI.Red.M.All.d , TypeI.Red.U.All.d , TypeI.Green.M.All.d , TypeI.Green.U.All.d , sample.call, marker.call, beta.raw, beta.raw.sex, pcaAuto, sample.callY, sample.callX,marker.call.all, marker.outlier.all,  file=paste0(outputdir,'/',qcfilename))

save(pcaControls, betaQN, betaQN.sex,betaQN.all,betaQN.all.3IQR, est.wbc.minfi, file=paste0(outputdir,"/",analysisfilename))
#save(pcaControls, betaQN, betaQN.sex,betaQN.all,betaQN.all.3IQR, mdata, mdata.3IQR, file=paste0(outputdir,"/",analysisfilename)) # with mdata 
```


```{r child = 'MemoryMethodsCredits.Rmd'}
```
