
## PART 4: Calculating Inter Quartile Ranges 

The inter-quartile range of a set of values is the difference between its 0.75-quartile and its 0.25-quartile. 
In the following, these quartiles are calculated for all autosomal beta values per probe and beta values whose difference to the median is greater than 3 times the inter-quartile range are marked as outliers and set to missing (NA) in betaQN.all.3IQR.

```{r}
  # outliers
  IQRrange <- 1.5
  betaQN.all.3IQR <- betaQN.all
  probe_IQR <- t(apply(betaQN.all,1,function(x) {quantile(x,probs=c(0.25,0.75,0.5),na.rm=T)}))
  probe_IQR <- data.frame(probe_IQR)
  probe_IQR$IQR <- probe_IQR[,2]-probe_IQR[,1]
  probe_IQR$rangeMin3IQR <- probe_IQR[,3]-3*probe_IQR$IQR # 3 IQR from median
  probe_IQR$rangeMax3IQR <- probe_IQR[,3]+3*probe_IQR$IQR # 3 IQR from median
  
  marker.outlier.all <- matrix(data=NA,ncol=3,nrow=nrow(betaQN.all.3IQR))
  cnt <- 0
  ## THIS TAKES VERY LONG !!! nrow(betaQN.all.3IQR)
   for (i in rownames(betaQN.all.3IQR)) {
       cnt <- cnt+1
       rangeMin <- probe_IQR[i,"rangeMin"]
       rangeMax <- probe_IQR[i,"rangeMax"]
       rangeMin3IQR <- probe_IQR[i,"rangeMin3IQR"]
       rangeMax3IQR <- probe_IQR[i,"rangeMax3IQR"]
   
       vals <- betaQN.all.3IQR[i,]
   
       nprobes<-length(which(!is.na(vals)))
       outlier3_IQR_vals <- which(vals<rangeMin3IQR | vals>rangeMax3IQR)
       outlier3_IQR<-length(outlier3_IQR_vals)
       
       betaQN.all.3IQR[i,outlier3_IQR_vals]<-NA
       marker.outlier.all[cnt,] <- c(i,nprobes,outlier3_IQR)
   
       #if (cnt %% 5000 == 0) { print(cnt) } #DEBUG
   }
   colnames(marker.outlier.all) <- c("name","nprobes","outlier3_IQR")
   marker.outlier.all <- as.data.frame(marker.outlier.all,stringsAsFactors=F)
   marker.outlier.all$nprobes<-as.numeric(marker.outlier.all$nprobes)
   marker.outlier.all$outlier3_IQR<-as.numeric(marker.outlier.all$outlier3_IQR)
   marker.outlier.all$callrate<-marker.outlier.all$nprobes/length(samples)
```


part                    | number of probes                                        | percentage 
------------------------|---------------------------------------------------------|-------------------------------------------------------------------------------------------
total                   | `r (nrow(betaQN.all)*ncol(betaQN.all))`                     | 100\%
noCall                  | `r length(betaQN.all[which(is.na(betaQN.all))])`            |`r length(betaQN.all[which(is.na(betaQN.all))])/((nrow(betaQN.all)*ncol(betaQN.all))) `
noCall with filter 3IQR | `r length(betaQN.all.3IQR[which(is.na(betaQN.all.3IQR))])`  | `r length(betaQN.all.3IQR[which(is.na(betaQN.all.3IQR))])/((nrow(betaQN.all)*ncol(betaQN.all))) `

```{r}   
   hist(marker.outlier.all$outlier3_IQR,breaks=50,main="#Probes Outlier 3 IQR",xlab="#outliers per probe",ylab="#probes")
```
  

```{r}
   hist(marker.outlier.all$nprobes,breaks=50,main="#Calls per Probe (QN Dataset)",xlab="#calls per probe",ylab="#probes")
```
  

```{r}
   hist(marker.outlier.all$callrate,breaks=50,main="#Callrate per Probe (QN Dataset)",xlab="callrate",ylab="#probes")
   
 # reformat output for analysis pipeline
 # convert array IDs to sample IDs
 ids <- datsamples[,c("Sample_ID","Sample_Name")]
 ids <- ids[match(colnames(betaQN.all),ids$Sample_ID),]

```
