
## PART 3: Quantile Normalisation
```{r}
###############################################################################

# PART 3 (QN)

# Perform Quantile Normalization based on final sample set
# Written by Benjamin Lehne (Imperial College London) and Alexander Drong (Oxford University)
# extended by Alexander Teumer (University Medicine Greifswald/ Erasmus MC Rotterdam)
# last change: 05/16/2016

datsamplesfinal <- read.table(file=samplesfilefinal,header=T,as.is=T,sep="\t")
samples <- colnames(TypeI.Red.M.All)
samples <- samples[samples %in% datsamplesfinal$Sentrix_Code]
datsamples <- datsamplesfinal[datsamplesfinal$Sentrix_Code %in% samples,]
```

`r nrow(datsamples)` samples are included in the analysis, identified by the *samplesfilefinal* `r samplesfilefinal ` and existing sample calls:

```{r}
(as.data.frame(datsamples))$Sample_Name
```

```{r}
#QN autosomes # has to be filtered by samples 
category=auto
markers=as.matrix(intersect(rownames(TypeII.Green.All.d), category))
TypeII.Green = TypeII.Green.All.d[markers,samples]
TypeII.Red = TypeII.Red.All.d[markers,samples]
markers=intersect(rownames(TypeI.Green.M.All.d), category)
TypeI.Green.M = TypeI.Green.M.All.d[markers,samples]
TypeI.Green.U = TypeI.Green.U.All.d[markers,samples]
markers=intersect(rownames(TypeI.Red.M.All.d), category)
TypeI.Red.M = TypeI.Red.M.All.d[markers,samples]
TypeI.Red.U = TypeI.Red.U.All.d[markers,samples]

TypeII.Green=normalizeQuantiles(TypeII.Green)
TypeII.Red = normalizeQuantiles(TypeII.Red)
TypeI.Green.M = normalizeQuantiles(TypeI.Green.M)
TypeI.Green.U = normalizeQuantiles(TypeI.Green.U)
TypeI.Red.M = normalizeQuantiles(TypeI.Red.M)
TypeI.Red.U = normalizeQuantiles(TypeI.Red.U)
TypeII.betas = TypeII.Green/(TypeII.Red+TypeII.Green+100)
TypeI.Green.betas = TypeI.Green.M/(TypeI.Green.M+TypeI.Green.U+100)
TypeI.Red.betas = TypeI.Red.M/(TypeI.Red.M+TypeI.Red.U+100)
betaQN = as.matrix(rbind(TypeII.betas,TypeI.Green.betas,TypeI.Red.betas))
marker.callQN =rowSums(!is.na(betaQN))/ncol(betaQN) # this line is added; if problems occur, maybe delete it and rechange all lines where marker.callQN occurs to marker.call

# sex chromosomes ------------------------------------------------------------------
women <- datsamples[datsamples$Gender==1,"Sentrix_Code"]
men <- datsamples[datsamples$Gender==0,"Sentrix_Code"]
category=c(sexchrX,sexchrY)

```

Of the persons from whom samples are processed, `r length(women)` are women and `r length(men)` men.   



```{r}
#QN men

markers=as.matrix(intersect(rownames(TypeII.Green.All.d), category))
TypeII.Green_m= TypeII.Green.All.d[markers,men]
TypeII.Red_m = TypeII.Red.All.d[markers,men]
markers=intersect(rownames(TypeI.Green.M.All.d), category)
TypeI.Green.M_m = TypeI.Green.M.All.d[markers,men]
TypeI.Green.U_m = TypeI.Green.U.All.d[markers,men]
markers=intersect(rownames(TypeI.Red.M.All.d), category)
TypeI.Red.M_m = TypeI.Red.M.All.d[markers,men]
TypeI.Red.U_m = TypeI.Red.U.All.d[markers,men]

# only if there are at least 2 men
if(length(men) > 1){
TypeII.Green_m=normalizeQuantiles(TypeII.Green_m)
TypeII.Red_m = normalizeQuantiles(TypeII.Red_m)
TypeI.Green.M_m = normalizeQuantiles(TypeI.Green.M_m)
TypeI.Green.U_m = normalizeQuantiles(TypeI.Green.U_m)
TypeI.Red.M_m = normalizeQuantiles(TypeI.Red.M_m)
TypeI.Red.U_m = normalizeQuantiles(TypeI.Red.U_m)
} else {
	cat("Quantile normalization for men was skipped, because it has no effect when there are less than 2 samples.")
}

TypeII.betas_m = TypeII.Green_m/(TypeII.Red_m+TypeII.Green_m+100)
TypeI.Green.betas_m = TypeI.Green.M_m/(TypeI.Green.M_m+TypeI.Green.U_m+100)
TypeI.Red.betas_m = TypeI.Red.M_m/(TypeI.Red.M_m+TypeI.Red.U_m+100)
beta_m = as.matrix(rbind(TypeII.betas_m,TypeI.Green.betas_m,TypeI.Red.betas_m))

#QN women
markers=as.matrix(intersect(rownames(TypeII.Green.All.d), category))
TypeII.Green_w= TypeII.Green.All.d[markers,women]
TypeII.Red_w = TypeII.Red.All.d[markers,women]
markers=intersect(rownames(TypeI.Green.M.All.d), category)
TypeI.Green.M_w = TypeI.Green.M.All.d[markers,women]
TypeI.Green.U_w = TypeI.Green.U.All.d[markers,women]
markers=intersect(rownames(TypeI.Red.M.All.d), category)
TypeI.Red.M_w = TypeI.Red.M.All.d[markers,women]
TypeI.Red.U_w = TypeI.Red.U.All.d[markers,women]


# only if there are at least 2 women
# watch out if you change this; this switch appears several times in the code due to formatting reasons
if(length(women) > 1){
TypeII.Green_w=normalizeQuantiles(TypeII.Green_w)
TypeII.Red_w = normalizeQuantiles(TypeII.Red_w)
TypeI.Green.M_w = normalizeQuantiles(TypeI.Green.M_w)
TypeI.Green.U_w = normalizeQuantiles(TypeI.Green.U_w)
 TypeI.Red.M_w = normalizeQuantiles(TypeI.Red.M_w)
 TypeI.Red.U_w = normalizeQuantiles(TypeI.Red.U_w)
} else {
	cat("Quantile normalization for women was skipped, because it has no effect when there are less than 2 samples.")
}

 TypeII.betas_w = TypeII.Green_w/(TypeII.Red_w+TypeII.Green_w+100)
TypeI.Green.betas_w = TypeI.Green.M_w/(TypeI.Green.M_w+TypeI.Green.U_w+100)
TypeI.Red.betas_w = TypeI.Red.M_w/(TypeI.Red.M_w+TypeI.Red.U_w+100)
beta_w = as.matrix(rbind(TypeII.betas_w,TypeI.Green.betas_w,TypeI.Red.betas_w))

betaQN.sex <- cbind(beta_m,beta_w) 
marker.call_sex=rowSums(!is.na(betaQN.sex))/ncol(betaQN.sex)
betaQN.all <- rbind(betaQN,betaQN.sex)

###### save data (combine with sex chr)

rm(TypeII.Green,TypeII.Red,TypeI.Green.M,TypeI.Green.U,TypeI.Red.M,TypeI.Red.U,TypeII.betas,TypeI.Green.betas,TypeI.Red.betas)
rm(TypeII.Green_m,TypeII.Red_m,TypeI.Green.M_m,TypeI.Green.U_m,TypeI.Red.M_m,TypeI.Red.U_m,TypeII.betas_m,TypeI.Green.betas_m,TypeI.Red.betas_m)
rm(TypeII.Green_w,TypeII.Red_w,TypeI.Green.M_w,TypeI.Green.U_w,TypeI.Red.M_w,TypeI.Red.U_w,TypeII.betas_w,TypeI.Green.betas_w,TypeI.Red.betas_w)

plot.beta.densities(betaQN,"Densities of normalized autosomal beta values per sample")
```

### PART 4: Control Probe Adjustment

Control probes track for technical bias between batches. To adjust for these in the analysis while minimizing the number of variables due to convergence reasons, we calculate Principal Components and show how much variance they explain. 
Markers which had at least one cpg site missing were excluded for the following PCA. 

```{r}
#PCA of control-probe intensities
datpca <- ctrl.all[samples,]
pcaControls <- prcomp(na.omit(datpca))
ctrlprobes.scores = pcaControls$x
EV <- merge(ctrlprobes.scores,datsamples,by.x=0,by.y="Sentrix_Code")
colnames(ctrlprobes.scores) = paste(colnames(ctrlprobes.scores), '_cp', sep='')

plot(EV[,"PC1"],EV[,"PC2"], main=paste0("PCA - number of controls probes: ",ncol(datpca)),xlab=paste0("PC1 (",sprintf("%1.1f",(pcaControls$sdev^2/sum(pcaControls$sdev^2))[1]*100),"%)"),ylab=paste0("PC2 (",sprintf("%1.1f",(pcaControls$sdev^2/sum(pcaControls$sdev^2))[2]*100),"%)"))

par(mfrow=c(1,1))
plot((pcaControls$sdev^2/sum(pcaControls$sdev^2))[1:10],main="Explained Variance by control probe PCs",ylab="% variance",xlab="PC#",ylim=c(0,1.05))
text((y=pcaControls$sdev^2/sum(pcaControls$sdev^2))[1:10],x=1:10+0.1,labels=sprintf("%1.3f",(pcaControls$sdev^2/sum(pcaControls$sdev^2)))[1:10],adj=c(0.5,-1))

ctrlprobes.scores <- merge(ctrlprobes.scores,datsamples[,c("Sentrix_Code","Sample_Name")],by.x=0,by.y="Sentrix_Code")
write.csv(ctrlprobes.scores,file=paste0(outputdir,"/PC_controls-finalSample","_",projectname,"-",formatted.time,".csv"),row.names=F)

# sex-check
sample.callY=colSums(!is.na(betaQN.sex[sexchrY,]))/nrow(betaQN.sex[sexchrY,])
sample.callX=colSums(!is.na(betaQN.sex[sexchrX,]))/nrow(betaQN.sex[sexchrX,])

# by PCA
```
  

The following plot is a PCA on **variables**: It presents the importance of samples when differentiating between the CpG sites in the best way possible. 

```{r}
# chr X+Y
betaPCA <- betaQN.sex[c(sexchrX,sexchrY),]
pcaSex <- prcomp(na.omit(betaPCA))
EV <- pcaSex$rotation
EV <- merge(EV,datsamples,by.x=0,by.y="Sentrix_Code")
plot(EV[,"PC1"],EV[,"PC2"],col=EV$Gender+2, main=paste0("number of X+Y probes: ",nrow(betaPCA)),xlab=paste0("EV1 (",sprintf("%1.1f",(pcaSex$sdev^2/sum(pcaSex$sdev^2))[1]*100),"%)"),ylab=paste0("EV2 (",sprintf("%1.1f",(pcaSex$sdev^2/sum(pcaSex$sdev^2))[2]*100),"%)"))
legend("top", c("male","female"), col=c(2,3), pch = 1, cex=1.4)

resSexCheck <- data.frame(Sentrix_Code=EV$Row.names,Sex=EV$Gender,EV1=EV[,"PC1"],EV2=EV[,"PC2"])
```
  


The next plot presents a PCA on **samples**, as it is more common. It shows the variance of the samples on the PC axes. 

```{r}
pcaSex <- prcomp(t(na.omit(betaPCA)))
EV <- pcaSex$x
EV <- merge(EV,datsamples,by.x=0,by.y="Sentrix_Code")
plot(EV[,"PC1"],EV[,"PC2"],col=EV$Gender+2, main=paste0("number of X+Y probes: ",nrow(betaPCA)),xlab=paste0("PC1 (",sprintf("%1.1f",(pcaSex$sdev^2/sum(pcaSex$sdev^2))[1]*100),"%)"),ylab=paste0("PC2 (",sprintf("%1.1f",(pcaSex$sdev^2/sum(pcaSex$sdev^2))[2]*100),"%)"))
legend("top", c("male","female"), col=c(2,3), pch = 1, cex=1.4)

resSexCheck <- cbind(resSexCheck,data.frame(PC1=EV[,"PC1"],PC2=EV[,"PC2"]))
write.csv(resSexCheck,file = paste0(outputdir,"/results-PCAsex-QN","_",projectname,"-",formatted.time,".csv"),row.names=F)

# plot sex call rates
YcallMen <- sample.callY[names(sample.callY) %in% men]
YcallWomen <- sample.callY[names(sample.callY) %in% women]
#par(mfrow=c(2,1))

if(length(men) > 1){
hist(YcallMen,breaks=50)
}
```
  



```{r}
if(length(women) > 1){
hist(YcallWomen,breaks=50)

marker.callYwomen=rowSums(!is.na(betaQN.sex[sexchrY,colnames(betaQN.sex) %in% women]))/ncol(betaQN.sex[sexchrY,colnames(betaQN.sex) %in% women])
}
```
  



```{r}
if(length(women) > 1){
par(mfrow=c(1,1))
hist(marker.callYwomen,breaks=50)
}
```

  

```{r}
# PCA (autosomes)
betaPCA <- betaQN
pcaAuto <- prcomp(na.omit(betaPCA))
EV <- pcaAuto$rotation
EV <- merge(EV,datsamples,by.x=0,by.y="Sentrix_Code")
par(mfrow=c(1,1))
plot(EV[,"PC1"],EV[,"PC2"], main=paste0("number of autosomal probes: ",nrow(pcaAuto$x)),xlab=paste0("EV1 (",sprintf("%1.1f",(pcaAuto$sdev^2/sum(pcaAuto$sdev^2))[1]*100),"%)"),ylab=paste0("EV2 (",sprintf("%1.1f",(pcaAuto$sdev^2/sum(pcaAuto$sdev^2))[2]*100),"%)"))

plot(EV[,"PC3"],EV[,"PC4"], main=paste0("number of autosomal probes: ",nrow(pcaAuto$x)),xlab=paste0("EV3 (",sprintf("%1.1f",(pcaAuto$sdev^2/sum(pcaAuto$sdev^2))[3]*100),"%)"),ylab=paste0("EV4 (",sprintf("%1.1f",(pcaAuto$sdev^2/sum(pcaAuto$sdev^2))[4]*100),"%)"))
```


```{r}
# EVals
par(mfrow=c(1,1))
plot((pcaAuto$sdev^2/sum(pcaAuto$sdev^2))[1:10],main="Explained Variance by autosomes",ylab="% variance",xlab="PC#",ylim=c(0,1.05))
text((y=pcaAuto$sdev^2/sum(pcaAuto$sdev^2))[1:10],x=1:10+0.1,labels=sprintf("%1.3f",(pcaAuto$sdev^2/sum(pcaAuto$sdev^2)))[1:10],adj=c(0.5,-1))

marker.call.all <- data.frame(callrate=c(marker.callQN,marker.call_sex)) # marker callrate of all samples incl. excluded ones
```

